|

### 2023-08-08 manuscript/paper.Rmd
Update manuscript

| Trait Name           | Trait Code | Trait Description           |
|----------------------|------------|-----------------------------|
|Standard Length|	sl|	Distance from anterior tip of premaxilla to posterior end of last vertebra (hypural plate)|
|Dorsal Spine | mds |	Number of dorsal spines from 0 to 3|
|Dorsal Fin Ray|mdf|	Number of bones in the dorsal fin posterior to the third dorsal spine (i.e., soft dorsal fin rays)
|Anal Fin Ray |maf|	Number of bones in the anal fin posterior to the anal spine (i.e., soft dorsal fin rays)|
|Abdominal Vertebra|mav|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Caudal Vertebra|mcv	| Number of vertebrae posterior to and including the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Pterygiophore number|mpt|	Number of pterygiophores anterior to but excluding the pterygiophore under the third dorsal spine, which is immediately anterior to and contiguous with the dorsal fin|
|Pelvic Spine length|lps.sc	|Length from the base of one pelvic spine above its articulation with the pelvic girdle to its distal tip|
|Ectocoracoid|ect.sc|	Length between the anterior and posterior tips of the shoulder girdle base (i.e., ectocoracoid)|
|Pelvic Girdle|tpg.sc|	Length between the anterior to posterior tips along midline. If vestigial, the sum of longest anterior-posterior axis for the vestiges|
|Cleithrum length|cle.sc|	Length from free dorsal tip to ventral tip of the cleithrum on the anterior margin of the shoulder girdle (i.e., cleithrum)|
|Premaxilla|pmx.sc|	Length from the anterior tip of the premaxilla to the distal tip of the ascending process of the premaxilla |
|Dorsal Spine|Ds#.sc# = 1,2,or 3|	Length from the base of a dorsal spine above the pterygiophore to its distal tip along the anterior edge|
|Pterygiophore|lpt.sc|	Distance between the anterior to posterior tips of the pterygiophore immediately preceding the 3rd dorsal spine (when present)|

@@

## Completed Data Model 
\textcolor{red}{We should note somewhere that we are only using the $W_{fossil}$ in the modeling part.  We drop the $W_{extant}$.  So just note that $W_{ti}$ is really $W_{fossil,ti}$.  Not sure how to say this, but we need to make it clear that we are only using the fossil data for the OU modeling.}

For a given imputed dataset, let $W_{ij}$ be the imputed gender and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fish $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$). In addition, we denote $Y_{K,ij}$, the last variable in $\boldsymbol{Y}_{ij}$, to be the standard length of the fish. We assume that $Y_{K,ij}$ follows
\begin{align}
{Y}_{K,ij} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right.
\label{eq:X}
\end{align}

<!-- ### Continuous Phenotypes -->
<!-- For a given imputed dataset, let $W_{ti}$ be the imputed gender, $\boldsymbol{X}_{ti}$ be the $K \times 1$ vector of continuous phenotypes, and $\boldsymbol{Y}_{ti}$ be the $L \times 1$ vector of discrete phenotypes for $t = 1, \ldots, T$ and  $i = 1\ldots,n_{t}$). -->

<!-- Define the last element of $\boldsymbol{X}_{ti}$ ($X_{K,ti}$) as the length of the stickleback.  For $k = 1,\cdots,K - 1$, we will assume -->
<!-- \begin{align} -->
<!-- {X}_{k,ti} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft} & + \beta_k(X_{K,ti} - \mu_{K,ft}),&\sigma_k^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{k,mt} & + \beta_k(X_{K,ti} - \mu_{K,mt}),&\sigma_k^2), & W_{ti} = \text{Male} \end{array}\right. \\ -->
<!-- {X}_{K,ti} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft},&\sigma_{K}^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{K,mt},&\sigma_{K}^2), & W_{ti} = \text{Male} \end{array}\right. -->
<!-- \label{eq:X} -->
<!-- \end{align} -->
<!-- where $\mu_{k,ft}$ and $\mu_{k,mt}$ represent the time-$t$ specific mean of phenotype $k$ for female and male stickleback fish, respectively, and $\beta_k$ is an additional parameter to account for the biological phenomenon that other continuous phenotypes for a particular animal are positively correlated with their length \textbf{CITATION NEEDED}.  We further set \begin{align} -->
<!-- \mu_{k,gt} = \theta_{k,g} + u_{k,gt}, -->
<!-- \label{eq:mu} -->
<!-- \end{align} -->
<!-- for $g \in \{f,m\}$ where $\theta_{k,g}$ is the overall mean of phenotype $X_k$ for each gender, and $u_{k,gt}$ measures the difference between $\mu_{k,gt}$ and $\theta_{k,g}$.  We then fit $\boldsymbol{u}_{k,g} = \{u_{k,g1},\cdots,u_{k,gT}\}$ to an Ornstein-Uhlenback (OU) process (@OUProcess) where we assume $u_{k,gT}$ have a marginal mean of 0. Because each time period is ~1000 years, we will discretize the OU process without loss of generality.  For $t = 2,\ldots,T$, we assume -->
<!-- \begin{align} -->
<!-- u_{k,gt} \iid \mathcal{N}(\kappa_{k} u_{k,g(t-1)} , \tau_k^2). -->
<!-- \label{eq:u_ar1} -->
<!-- \end{align} -->

<!-- $\kappa_{k,g}$ represents the correlation between $\mu_{k,gt}$ and $\mu_{k,g(t+1)}$, and it also follows that $\text{cor}(u_{g,t},u_{g,t+h}) = \kappa^h$ for $h \in \mathbb{N}$.  In addition, we assume -->
<!-- \begin{align} -->
<!-- u_{k,g1} \iid \mathcal{N}\left(0,\frac{\tau_k^2}{1-\kappa_{k}^2}\right). -->
<!-- \label{eq:u1} -->
<!-- \end{align} -->
<!-- This model choice is to preserve stationarity; i.e. $p(u_{k,gs}) = p(u_{k,gt})$ for $s \neq t$. -->

<!-- ### Discrete Phenotypes -->
<!-- For the discrete phenotypes, we will assume -->
<!-- \begin{align} -->
<!-- {Y}_{l,ti} & \sim \left\{\begin{array}{ll} Poisson(\lambda_{l,ft}), & W_{ti} = \text{Female} \\ Poisson(\lambda_{l,mt}), & W_{ti} = \text{Male} \end{array}\right., -->
<!-- \label{eq:Y} -->
<!-- \end{align} -->
<!-- for $l = 1,\cdots,L$.  Similar to the continuous phenotypes, we set -->
<!-- \begin{align} -->
<!-- \log(\lambda_{l,gt}) = \gamma_{l,g} + v_{l,gt}. -->
<!-- \label{eq:lambda} -->
<!-- \end{align} -->
<!-- We use the log function to allow $v_{l,gt}$ to take all values on the real number line so we can properly fit an OU process to these effects.  We further assume -->
<!-- \begin{align} -->
<!-- v_{l,gt} \iid \mathcal{N}(\phi_{l} v_{l,g(t-1)} , \omega_l^2), -->
<!-- \label{eq:v_ar1} -->
<!-- \end{align} -->
<!-- and -->
<!-- \begin{align} -->
<!-- v_{l,g1} \iid \mathcal{N}\left(0,\frac{\omega_l^2}{1 - \phi_{l}^2}\right).. -->
<!-- \label{eq:v1} -->
<!-- \end{align} -->

@@

Priors: For $k = 1,\cdots,K$ and $l = 1,\cdots,L$,

\begin{align}
\sigma_k & \iid \mathcal{N}(0,100)I_{\{\sigma > 0\}} \nonumber \\
\tau_k & \iid \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
\kappa_k & \iid \mathcal{N}(0,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
\beta_k & \iid \mathcal{N}(0,5) \nonumber \\
\theta_{k,g} & \iid \mathcal{N}(0,10000) \nonumber \\
\omega_l & \iid \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
\phi_l & \iid \mathcal{N}(0,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
\gamma_{l,g} & \iid \mathcal{N}(0,10000)
\label{eq:priors}
\end{align}

All models were built using @R2022language

>>>>

| Trait Name           | Trait Code | Trait Description           |
|----------------------|------------|-----------------------------|
|Standard Length|	stl|	Distance from anterior tip of premaxilla to posterior end of last vertebra (hypural plate)|
|Dorsal Spine | mds |	Number of dorsal spines from 0 to 3|
|Dorsal Fin Ray|mdf|	Number of bones in the dorsal fin posterior to the third dorsal spine (i.e., soft dorsal fin rays)
|Anal Fin Ray |maf|	Number of bones in the anal fin posterior to the anal spine (i.e., soft dorsal fin rays)|
|Abdominal Vertebra|mav|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Caudal Vertebra|mcv	| Number of vertebrae posterior to and including the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Pterygiophore number|mpt|	Number of pterygiophores anterior to but excluding the pterygiophore under the third dorsal spine, which is immediately anterior to and contiguous with the dorsal fin|
|Pelvic Spine length|lps.sc	|Length from the base of one pelvic spine above its articulation with the pelvic girdle to its distal tip|
|Ectocoracoid|ect.sc|	Length between the anterior and posterior tips of the shoulder girdle base (i.e., ectocoracoid)|
|Pelvic Girdle|tpg.sc|	Length between the anterior to posterior tips along midline. If vestigial, the sum of longest anterior-posterior axis for the vestiges|
|Cleithrum length|cle.sc|	Length from free dorsal tip to ventral tip of the cleithrum on the anterior margin of the shoulder girdle (i.e., cleithrum)|
|Premaxilla|pmx.sc|	Length from the anterior tip of the premaxilla to the distal tip of the ascending process of the premaxilla |
|Dorsal Spine|Ds#.sc# = 1,2,or 3|	Length from the base of a dorsal spine above the pterygiophore to its distal tip along the anterior edge|
|Pterygiophore|lpt.sc|	Distance between the anterior to posterior tips of the pterygiophore immediately preceding the 3rd dorsal spine (when present)|

@@

## Completed Data Model 
\textcolor{red}{We should note somewhere that we are only using the $W_{fossil}$ in the modeling part.  We drop the $W_{extant}$.  So just note that $W_{ti}$ is really $W_{fossil,ti}$.  Not sure how to say this, but we need to make it clear that we are only using the fossil data for the OU modeling.  Greg, check and see if this makes sense.}

For a given imputed dataset, let $W_{ij}$ be the imputed gender and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fossil $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$. Note, we are inputting only the fossil data into our model; connecting to the previous section, $W_{ij}$ and $\boldsymbol{Y}_{ij}$ can be interpretted as $W_{fossil,ij}$ and $\boldsymbol{Y}_{fossil,ij}$. In addition, we denote $Y_{K,ij}$, the last variable in $\boldsymbol{Y}_{ij}$, to be the standard length of the fish,
\begin{align}
{Y}_{K,ij} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:stl}
\end{align}

It is reasonable to assume that the other continuous traits of stickleback fish will have some correlation with its standard length \textbf{(CITATION)}. We account for this by adding an additional parameter, $\gamma_k$, into our model. More specifically, if $Y_{k,ij}$ is a continuous trait, then
\begin{align}
{Y}_{k,ij} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:cont}
\end{align}

In addition, we assume that the discrete traits abdominal vertebrae (mav) and caudal vertebrae (mcv) also have correlation with the standard length of the fish (\textbf{CITATION}). If $Y_{k,ij}$ is one of the above traits, then
\begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} Poisson(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\}), & W_{ij} = \text{Female} \\ Poisson(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\}), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_corr}
\end{align}

For the other discrete traits, we also assume the above model except we set $\gamma_k = 0$ because of the assumption of no correlation between the standard length and these traits. More specifically, we assume
\begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} Poisson(\exp\{\mu_{k,ft_i}\}), & W_{ij} = \text{Female} \\ Poisson(\exp\{\mu_{k,mt_i}\}), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_ind}
\end{align}

In the above model descriptions, $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$ model the time-$t_i$ specific mean of phenotype $k$ for female and male stickleback fish, respectively. We point out that, for the discrete traits, the means are represented by $\exp\{\mu_{k,ft_i}\}$ and $\exp\{\mu_{k,mt_i}\}$ for ease of use in our modeling technique. We further set 
\begin{align}
\mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
\label{eq:mu}
\end{align}
for $g \in \{f,m\}$ where $\beta_{0,kg}$ and $\beta_{1,kg}$ are regression parameters of phenotype $Y_k$ for each gender, accounting for the possibility of a time-dependent trend in the mean structure, and $u_{k,gt_i}$ is the corresponding residual. To account for potential correlations between the residuals for a given trait $k$ and gender $g$, we fit an Ornstein-Uhlenback (OU) process (@OUProcess). More specifically, define $du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}$, the change in $u_k,gt$ for a given trait $k$ and gender $g$ over a miniscule time period $dt$. The OU process is defined as 
\begin{align}
du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
\label{eq:OU_cont}
\end{align}
where $\kappa_k$ is a parameter associated with the correlation between $u_{k,gt}$ and $u_{k,g(t+dt)}$, $\tau_k$ is the standard deviation of the OU process, and $W_t$ is a standard Brownian motion. As shown in (@OUProcess), the closed form solution for the SDE in (\ref{eq:OU_cont}) is
\begin{align}
u_{k,gt_i} \iid \mathcal{N}\left(u_{k,gt_{i-1}}\exp\{-\kappa_k(t_{i} - t_{i-1})\} , \frac{\tau_k^2(1 - \exp\{-2\kappa_k(t_{i} - t_{i-1}))\}}{2\kappa_k}\right)
\label{eq:OU_sol}
\end{align}
for $i = 2,\cdots,T$. In a traditional OU process, the initial value $u_{k,gt_1}$ is assumed to be a (potentially unknown) constant, and we discuss the procedure for estimating this value below.

In our empirical study, we analyze the following four nested models for the mean process outlined in (\ref{eq:mu}):
\begin{itemize}
\item OU with Trend: The mean process for each phenotype $k=1,\cdots,K$ and $g \in \{f,m\}$ as previously defined.
\item OU with No Trend: No linear trend on the mean processes and the overall mean of the phenotype is constant for each gender. This is achieved by setting $\beta_{1,kg} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$ and $g \in \{f,m\}$
\item No OU with Trend: No random fluctuations on the mean processes, i.e. the OU process assumption is not needed. This is achieved by setting $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
\item No OU with No Trend: The mean for a specific trait and specific gender at each time point is a constant value. This is achieved by setting $\beta_{1,kg} = 0$ and $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
\end{itemize}

@@

Priors: For $k = 1,\cdots,K$,

\begin{align}
u_{k,gt_1} & \iid \mathcal{N}(0,\tau_{0,k}) \nonumber \\
\sigma_k & \iid \mathcal{N}(0,10)I_{\{\sigma > 0\}} \nonumber \\
\tau_k & \iid \mathcal{N}(0,10)I_{\{\tau > 0\}} \nonumber \\
\tau_{0,k} & \iid \mathcal{N}(0,20)I_{\{\tau > 0\}} \nonumber \\
\kappa_k & \iid \mathcal{N}(0,1)I_{\{\kappa_g > 0\}} \nonumber \\
\gamma_{k} & \iid \mathcal{N}(0,5) \nonumber \\
\beta_{0,kg} & \iid \mathcal{N}(0,100) \nonumber \\
\beta_{1,kg} & \iid \mathcal{N}(0,3) \nonumber \\
\label{eq:priors}
\end{align}
We also note that, for the discrete phenotypes in equations (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}), there is no $\sigma_k$, and thus is not sampled for those phenotypes.

All models were built using @R2022language and (\textbf STAN)
